---
title: "R : Exploring Data for Machine Learning Model"
author: "Stephanie W"
output: html_document
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
library(knitr)
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, fig.align = 'center', dpi = 100, tidy = T, cache = F, cache.path = '.cache/', fig.path = 'fig/')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

When exploring data for Machine Learning, we're looking for:

- imbalance outcomes/predictors
- outliners
- groups of outcome points not explained by any of the predictors
- skewed variables (that needs to be transformed)

We'll use the Wage dataset from the ISLR package.  
This dataset reports wage and other data (age, education, jobclass, etc.) for a group of 3000 male workers in the Mid-Atlantic region.
```{r}
library(ISLR)
data(Wage)
summary(Wage)
```
Building training and testing sets (50% of the dataset each):
```{r}
library(caret)
intrain <- createDataPartition(y=Wage$wage, p=0.5, list =F)
training = Wage[intrain,]
testing = Wage[-intrain,]
```

The exploration is always done on the training set.

## Plotting predictors against outcome

Plotting wage versus age, education and jobclass using the R featurePlot function (from the caret package):
```{r}
featurePlot(x=training[,c("age","education", "jobclass")], y=training$wage, plot="pairs")
```

Plotting wage versus age:
```{r}
library(ggplot2)
qplot(age, wage, data=training)
```
The graph shows some patterns: a trend in wages comparing to ages and two distinct groups of observations (below and above 250 dollars raw wage).

Plotting wage versus age grouping by jobclass:
```{r}
qplot(age, wage, color=jobclass, data=training)
```
The jobclass difference could explain the two distincts groups.

Plotting wage versus age, grouping by education, adding regression smoothers:
```{r}
qq <- qplot(age, wage, color=education, data=training)
qq + geom_smooth(method="lm", formula=y~x)
```

## Data Repartition

Breaking up the wage variable into three groups (factors actually) with the R cut2 function (from the Hmisc package):
```{r}
library(Hmisc)
cutWage <- cut2(training$wage, g=3)
table(cutWage)
```

Looling at the repartition, we can see that there are more industrial jobs that there are information jobs with lower wage. Then the trend reverses itself.

Plotting a boxplot of the wage groups created above:
```{r}
qplot(cutWage, age, data=training, fill=cutWage, geom=c("boxplot"))
```

Exploring the repartition of jobclass across wage groups:
```{r}
t1 <- table(cutWage, training$jobclass)
t1
```

Using the prop function to get the proportion of jobclass (in each row) for each groups:
```{r}
prop.table(t1, 1)
```
62% of the low wage job correponds to industrial jobs, 38% to information jobs.

## Density Plots
 
Density plot can be a much more effective way to view the distribution of a variable than boxplots.

Ploting a density plot of the values of wages grouping by education:
```{r}
qplot(wage, color=education, data=training, geom="density")
```
The "<HS grad" workers tend to have more values in the lower part of the range of wage. There is an outgroup of Advanced Degree and College Grad workers with higher wage.
