---
title: "Earthquakes from the past 30 days"
output:
  html_document:
    theme: flatly
    highlight: tango
    
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
library(knitr)
options(options(reindent.spaces = 2))
opts_chunk$set(message = F, error = F, warning = F, fig.align = 'center', dpi = 100, tidy = F, cache = F, cache.path = '.cache/', fig.width=8, fig.path = 'figure/')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)){ 
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


Earthquakes of the last 30 months from the earthquake.usgs.gov feed : http://earthquake.usgs.gov/earthquakes/feed/v1.0/csv.php feed. 


Below are the fields included in the spreadsheet output:

    time
    latitude
    longitude
    depth
    mag
    magType
    nst
    gap
    dmin
    rms
    net
    id
    updated
    place

Loading libraries and downloading the feed:
```{r}
library(reshape2)
library(ggplot2)
library(ggmap)
eq <- read.csv("http://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv", as.is = T)
head(eq[c("time", "longitude", "latitude", "place", "mag")])
```

Preprocessing the data:
```{r}
eq <- subset(eq, type=="earthquake")
eq$area <- factor(sub("^[^,]+, ", "", eq$place))
eq$date <- as.Date(strtrim(eq$time, 19), format = "%Y-%m-%dT%H:%M:%S")
```

Building a new dataset with magnitude frequencies by day:
```{r}
eqFreq1 <- with(eq, table(date, mag))
eqFreq2 <- melt(eqFreq1)
names(eqFreq2) <- c("date", "magnitude", "freq")
head(subset(eqFreq2, freq > 0 & magnitude > 0))
```

Plotting a stacked bars graph of earthquakes magnitude frequencies by day:
```{r earthquakes_frequency}
eqFreq2$magnitude <- factor(round(eqFreq2$magnitude))
ggplot(eqFreq2, aes(date, weight = freq, fill = magnitude)) +
        geom_bar(binwidth = 60 * 60 * 24) +
        labs(x = "Date", y = "Frequency", 
             title = "Earthquakes Frequency from the past 30 days") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        scale_fill_brewer(palette="BuPu")
```

Using the original dataset to plot a map with the locations of the 100 biggest earthquakes:
```{r earthquakes_worldmap}
BiggestMag <- tail(sort(eq$mag), n=100)
#Plotting
library(maps)
world_map <- map_data("world")
ggplot() + coord_fixed() + xlab("") + ylab("") +
        geom_polygon(data = world_map, 
                     aes(x = long, y = lat, group = group), 
                     colour = "light green", fill = "light green") +
        geom_point(aes(x = longitude, y = latitude, size = mag),
                   data = eq[eq$mag %in% BiggestMag ,], 
                   colour = "Deep Pink", fill = "Pink", 
                   pch = 21, alpha = I(0.7)) +
        labs(title="Biggest earthquakes from the past 30 days")
```
